# Prepare data

## Sample metadata
```{r}

```

## Genome metadata
We are preparing a table with ou MAG catalogue information, by combining taxonomy (from GTDB-Tk) with genome characteristics.
```{r prepare_genome_metadata, warning=FALSE, comments="", message=FALSE}
# Load genome information and clean it
genome_info <- read_csv("resources/genomeInfo.csv") %>% # Read genome info file where you can find the completeness,contamination and length of our MAGs
  select(-N50) %>% # Drop N50 column (not needed for downstream analysis here)
  mutate(genome = str_replace(genome, "\\.fa$", "")) # Remove ".fa" suffix from genome names for consistency

# Load GTDB-Tk genome taxonomy summary
genome_metadata <- read_tsv("resources/gtdbtk.bac120.summary.tsv")%>%
  dplyr::rename(genome=1) %>% # Rename first column to "genome" to match genome_info
  mutate(classification = str_replace_all(classification, ".__", "")) %>% # Remove prefix from taxonomy labels
  separate(col = classification, sep = ";", into = c("domain", "phylum", "class", "order", "family", "genus", "species")) %>% # Split taxonomy string into individual taxonomic levels
  select(1:8) %>% # Keep genome + taxonomy columns
  left_join(., genome_info, by="genome") # Add additional genome info (completeness,contamination and length) from genome_info
```

## Read counts
This table contains the number of sequencing reads that map to each genome per sample. It’s the raw input for abundance calculations.
```{r load_reads, warning=FALSE, comments="", message=FALSE}
# Load raw read counts for each genome in each sample
counts_raw <- read_tsv("resources/coverm_read_counts.tsv")
```

## Genome counts
```{r prepare_genome_counts, warning=FALSE, comments="", message=FALSE}
min_coverage = 0.3 # Minimum fraction of genome coverage to consider the genome "present" in a sample

# Load covered bases per genome and convert to counts. Ensures only sufficiently covered genomes are considered to reduce noise from spurious reads
read_counts_filt <- read_tsv("resources/coverm_covered_bases.tsv") %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>% # Apply coverage threshold: 1 if genome sufficiently covered, 0 if not
  arrange(match(Genome,counts_raw$Genome)) %>% # Ensure genomes are in the same order as counts_raw
  mutate(across(-1, ~ . * counts_raw[[cur_column()]])) # Multiply presence/absence by raw read counts

# Normalise genome counts by genome length. Dividing by genome length avoids bias from large genomes naturally having more reads.
genome_counts_filt <- read_counts_filt %>%
  arrange(match(Genome,genome_metadata$genome)) %>% 
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / read_length) ))%>%
  select_if(~ !is.numeric(.) || sum(.) != 0)%>% #Keeps samples where the sum of reads across all MAGs is greater than 0.
  filter(rowSums(across(where(is.numeric)))!=0) #Keeps rows where the sum of reads across all samples is greater than 0.

genome_metadata <- genome_metadata %>% 
  filter(genome %in% genome_counts_filt$Genome) # Ensure genomes filtered out from genome_counts_filt are also removed from genome_metadata
```

## Genome tree
Phylogenetic trees allow us to visualize evolutionary relationships among genomes. Keeping only relevant MAGs ensures consistency between tree and abundance/metadata tables.
```{r load_genome_tree, warning=FALSE, comments="", message=FALSE}
# Load genome phylogenetic tree
genome_tree <- read_tree("resources/gtdbtk.backbone.bac120.classify.tree")

# Keep only the tips corresponding to MAGs in our dataset
genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips
```

## Prepare color scheme
This allows consistent coloring of taxa in visualizations, making plots easier to interpret.
```{r get_ehi_colors_skin, warning=FALSE, comments="", message=FALSE}
phylum_colors <- read_tsv(
  "https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", "")) %>% # Clean taxonomy prefix
  right_join(genome_metadata, by = join_by(phylum == phylum)) %>% # Keep only phyla in our dataset
  arrange(match(genome, genome_tree$tip.label)) %>% # Align colors with tree tips
  select(phylum, colors) %>%
  unique() %>%
  arrange(phylum) %>%
  pull(colors, name = phylum)
```

# Wrap working objects
All working objects are wrapped into a single Rdata object to facilitate downstream usage.
```{r savedata, comment="", message=FALSE, warning=FALSE}
save(sample_metadata,
  genome_metadata,
  genome_counts_filt,
#  preprocess_info,
  genome_tree,
#  genome_gifts,
#  microbial_fraction,
  phylum_colors,
  file = "resources/metagenomics/data.Rdata")
```
Saving as .Rdata is convenient for large projects: all processed data can be loaded quickly without repeating preprocessing steps. It also ensures reproducibility.


